A Convolutional Neural Network for Outsole Recognition\tnoteref{t1}

### 2019-06-07 paper/paper.Rnw
Get rid of floatbarrier - it won't compile with that in there.
comments

In this paper, we discuss an alternate approach that uses an additional model layer to transfer the feature vector output into a vector of probabilities representing the detection of geometric elements in shoe tread images. Working within the feature space used by forensic examiners allows us to augment human-identified features with model output, assessing similarity of different shoe images on a feature set that is explicitly relevant to the domain.
%\svp{Add 1-2 sentences connecting feature sets to the goal of reducing workload - possibly cite solemate as an example of manual feature identification}
\mt{In addition, much of the existing software to assess geometric similarity, such as SoleMate \fix{(how to cite software?)}, requires examiners to identify and label features manually; automating classification within the currently used feature space could reduce or eliminate the need for such manual classification.}

\section{Materials and Methods}
%\svp{Short bit (1-2 sentences) about low-level feature detectors and why they fail with this problem.}
\mt{Shoeprint evidence from crime scenes is most commonly collected in the form of a photograph, so any useful method to automatically identify outsole characteristics must take the form of an image analysis task. There are a number of methods that may be employed to identify shapes and features in an image, such as generalized Hough transformations \citep{ballardGeneralizingHoughTransform1981} and other low-level feature extraction methods aimed at detecting specific shapes, such as edges, corners, blobs, or ridges \citep[Ch 15]{machineVision}. While these methods are useful in identifying these specific features at a low level, they are very computationally intensive and only identify features on a very small scale; as a result, they cannot reliably identify large geometric shapes like those that may be found in an outsole image.} \fix{Necessary?: In addition, as one or more models would be required for each geometric shape, classical methods would require use of additional machine learning models such as random forests to aggregate low-level features into functional geometries found on outsoles.}

%\svp{Motivate the use of CNNs directly - complex model, higher level of feature aggregation. Mention pre-trained networks to transition.}
Convolutional neural networks (CNNs) are widely recognized as superior for novel image classification. CNNs are a form of artificial neural network which make use of the image convolution operator used by many low-level feature extraction methods, \mt{with the additional ability to aggregate such features and meaningfully connect them to a pre-determined set of labels}. CNNs have deep architectures that can be trained to identify complex patterns, but they are structurally similar to the human visual architecture and output binary or probabilistic predictions for given labels that are readily interpretable. As CNNs make use of labeled training data, the predictions generated are for features which are similar to those identified by humans, resulting in models with greater face validity. Once a CNN is trained, it is relatively fast and easy to apply the model to new images and obtain classifications.

@@

Many convolutional neural networks have been trained on a set of images known as ImageNet; these images are hierarchically labeled, so the same image of a dog may have labels of sky, grass, and trees, and the dog may be labeled as ``mammal'', ``dog'', ``retriever'', and ``golden retriever'' at increasing levels of label complexity \citep{dengImageNetLargeScaleHierarchical}. CNNs trained on ImageNet are optimized for general human-like vision, that is, the ability to recognize a large set of different features simultaneously. As a result, the base of networks trained on these sets are often used for transfer learning, because the initial layers are broadly generalizable to a wide variety of more specific image labeling tasks. \autoref{fig:vgg16layers} shows a selection of filters from VGG16, where each filter detects the presence of specific shapes and colors. \mt{Each row of filters in \autoref{fig:vgg16layers} corresponds to one of the five convolutional blocks of VGG16.} Filters in earlier convolutional blocks detect simple features, such as colors and blobs, while later filters detect more complex combinations of features. Each of the five convolutional blocks has between 128 and 1,536 filters, and a single image is made up of some combination of the thousands of features that are detected by those filters.

\mt{Each layer in the model structure is comprised of sets of weights and connections that have been optimized during training.} Transfer learning in neural network contexts involves freezing the weights of the pre-trained model base taken from a network trained on a more general set of images, so that during the training process, only the weights of the newly-added classifier are updated \citep{oquabLearningTransferringMidlevel2014}. In many cases, the entire model base is used, and fitted with a new model head; the modularity of CNNs makes this process relatively simple, and allows researchers to leverage pre-trained networks when working with diffierent sets of image data. Transfer learning allows CNNs to be applied to smaller datasets of several thousand images, reduces the amount of computational time required to fit the model, and provides boosted performance compared to training a new model from scratch \citep{yosinskiHowTransferableAre2014}. It has been successfully applied in automatic classification of medical images \citep{shinDeepConvolutionalNeural2016} as well as in various applications of shoe forensics \citep{kongCrossDomainForensicShoeprint2017,kongCrossDomainImageMatching2019a,zhangAdaptingConvolutionalNeural2017}.

In this paper, we will demonstrate the use of transfer learning to automatically identify features in shoe treads which are used by forensic examiners, using a general-purpose CNN with relatively simple structure, VGG16, and a database of labeled shoe tread images we have assembled for this project, which are both described below. This approach differs from other approaches \citep{kongCrossDomainForensicShoeprint2017,kongCrossDomainImageMatching2019a,zhangAdaptingConvolutionalNeural2017} in automatic footwear identification which use the output from the model base directly and do not attempt to add human-friendly contextual information with an additional classifier.

@@

<<vgg16layers, fig.width = 7, fig.height = 5.75, fig.cap = "A selection of filters from the convolutional layers of VGG16.">>=
imgs <- list.files("./images/vgg16layers/", recursive = T, full.names = T)
pngs <- lapply(imgs, function(x) png::readPNG(x)[46:1596, 32:1582, ])
grobs <- lapply(pngs, grid::rasterGrob)
gridExtra::grid.arrange(grobs = grobs, nrow = 5)
@

@@

The categories we use in this study are operationally defined as follows:
\begin{description}
\item [Bowtie] Bowtie shapes are roughly quadrilateral, with two opposite concave faces. The remaining two faces can be convex or straight, and the concave faces may have straight portions, so long as there is a concave region. Using this definition, shapes such as butterflies are included as bowties. \svp{I'm still on the fence - we can just not mention this, or we could move butterflies to "star"...} \mt{I think it makes a lot of sense to move them to star. That means redoing the poster/slides examples though... grrrrrr}
\item [Chevron] Chevron shapes include repeating parallel lines as well as individual ``v" shapes. They may be angular but can also be curved.
\item [Circle] Circles include ellipses and ovals; they must be round.
\item [Line] Lines are repeated and parallel; a more general definition of a line would be difficult to differentiate from many other patterns. Lines can be mildly curved.
\item [Polygon] Polygons are defined in this standard to have more than 4 sides. They include pentagons, hexagons, and octagons.
\item [Quadrilateral] Quadrilaterals (quads) have four sides. They may have rounded or square corners.
\item [Star] Stars are any shape with alternating concave and convex regions, or lines which emanate from a central point. ``X" and ``+" shapes are also classified as stars.
\item [Text] Text is any shape which would be identified as text by a reasonable human. In most cases, the text on the outsole images used is made up of Latin alphabet characters; the model will likely not recognize text in other scripts (but could be trained if non-Latin text images could be obtained).
\item [Triangle] Triangles are any three-sided figure. Like quadrilaterals, they can have rounded corners. In some cases, it is difficult to distinguish between a trapezoidal shape and a triangle when rounded corners are involved.
\item [Other] Other features which were marked include logos, various textures (including crepe, stippling, etc.), and smooth regions with no discernible features. These regions are grouped and provide additional information---that none of the previous nine categories are present.
\end{description}

@@

\FloatBarrier

>>>>

In this paper, we discuss an alternate approach that uses an additional model layer to transfer the feature vector output into a vector of probabilities representing the detection of geometric elements in shoe tread images. Working within the feature space used by forensic examiners allows us to augment human-identified features with model output, assessing similarity of different shoe images on a feature set that is explicitly relevant to the domain.
%\svp{Add 1-2 sentences connecting feature sets to the goal of reducing workload - possibly cite solemate as an example of manual feature identification}
In addition, much of the existing software to assess geometric similarity, such as SoleMate\footnote{\url{http://www.fosterfreeman.com/trace-evidence/357-sicar-6-solemate-2.html}}, requires examiners to identify and label features manually; automating classification within the currently used feature space could reduce or eliminate the need for such manual classification.

\section{Materials and Methods}
%\svp{Short bit (1-2 sentences) about low-level feature detectors and why they fail with this problem.}
\mt{Shoeprint evidence from crime scenes is most commonly collected in the form of a photograph, so any useful method to automatically identify outsole characteristics must take the form of an image analysis task.}\svp{We're not working with prints, though - please try to avoid the temptation to delve too far into prints. "Ultimately, most pattern evidence is stored in photographic form; as a result, we must use image analysis methods to create numerical features from the original visual record." or something to that effect.}
There are a number of methods that may be employed to identify shapes and features in an image, such as \svp{Fourier transforms, }Hough transformations \citep{ballardGeneralizingHoughTransform1981} and other low-level feature extraction methods aimed at detecting specific shapes, such as edges, corners, blobs, or ridges \citep[Ch 15]{machineVision}. While these methods are useful in identifying these specific features at a low level, they can be computationally intensive and only identify features on a very small scale; as a result, they cannot reliably identify large geometric shapes like those that may be found in an outsole image.
%\svp{Motivate the use of CNNs directly - complex model, higher level of feature aggregation. Mention pre-trained networks to transition.}
Convolutional neural networks (CNNs) are widely recognized as superior for novel image classification. CNNs are a form of artificial neural network which make use of the image convolution operator used by many low-level feature extraction methods, with the additional ability to aggregate such features and meaningfully connect them to a pre-determined set of labels. CNNs have deep architectures that can be trained to identify complex patterns, but they are structurally similar to the \svp{architecture of the }human visual \svp{system} and output binary or probabilistic predictions for given labels that are readily interpretable. As CNNs make use of labeled training data, the predictions generated are for features which are similar to those identified by humans, resulting in models with greater face validity. Once a CNN is trained, it is relatively fast and easy to apply the model to new images and obtain classifications.

@@

Many convolutional neural networks have been trained on a set of images known as ImageNet; these images are hierarchically labeled, so the same image of a dog may have labels of sky, grass, and trees, and the dog may be labeled as ``mammal'', ``dog'', ``retriever'', and ``golden retriever'' at increasing levels of label complexity \citep{dengImageNetLargeScaleHierarchical}. CNNs trained on ImageNet are optimized for general human-like vision, that is, the ability to recognize a large set of different features simultaneously. As a result, the base of networks trained on these sets are often used for transfer learning, because the initial layers are broadly generalizable to a wide variety of more specific image labeling tasks. \svp{You need to talk about the model structure of vgg16 before you start talking about layers. Talk of convolutional blocks doesn't really make sense until you show the layer structure.}\autoref{fig:vgg16layers} shows a selection of filters from VGG16, where each filter detects the presence of specific shapes and colors. Each row of filters in \autoref{fig:vgg16layers} corresponds to one of the five convolutional blocks of VGG16. Filters in earlier convolutional blocks detect simple features, such as colors and blobs, while later filters detect more complex combinations of features. Each of the five convolutional blocks has between 128 and 1,536 filters, and a single image is made up of some combination of the thousands of features that are detected by those filters.

Each layer in the model structure is comprised of sets of weights and connections that have been optimized during training. Transfer learning in neural network contexts involves freezing the weights of the pre-trained model base taken from a network trained on a more general set of images, so that during the training process, only the weights of the newly-added classifier are updated \citep{oquabLearningTransferringMidlevel2014}. In many cases, the entire model base is used, and fitted with a new model head; the modularity of CNNs makes this process relatively simple, and allows researchers to leverage pre-trained networks when working with diffierent sets of image data. Transfer learning allows CNNs to be applied to smaller datasets of several thousand images, reduces the amount of computational time required to fit the model, and provides boosted performance compared to training a new model from scratch \citep{yosinskiHowTransferableAre2014}. It has been successfully applied in automatic classification of medical images \citep{shinDeepConvolutionalNeural2016} as well as in various applications of shoe forensics \citep{kongCrossDomainForensicShoeprint2017,kongCrossDomainImageMatching2019a,zhangAdaptingConvolutionalNeural2017}.

\fix{In this paper, we will use transfer learning to automatically identify features in shoe treads which are used by forensic examiners, leveraging a general-purpose CNN with relatively simple structure, VGG16, and a database of labeled shoe tread images we have assembled for this project, which are both described below.} \svp{This sentence is a bit awkward, because you're introducing too many things. The paragraph is good, but I don't think that it makes sense to have it here. Introduce VGG16, then use this paragraph as a transition to talking about the actual model we fit.} This approach differs from other approaches \citep{kongCrossDomainForensicShoeprint2017,kongCrossDomainImageMatching2019a,zhangAdaptingConvolutionalNeural2017} in automatic footwear identification which use the output from the model base directly and do not attempt to add human-friendly contextual information with an additional classifier.

@@

<<vgg16layers, fig.width = 7, fig.height = 5.75, fig.cap = "A selection of filters from the convolutional layers of VGG16.\\svp{Label the rows with block numbers please}">>=
imgs <- list.files("./images/vgg16layers/", recursive = T, full.names = T)
pngs <- lapply(imgs, function(x) png::readPNG(x)[46:1596, 32:1582, ])
grobs <- lapply(pngs, grid::rasterGrob)
gridExtra::grid.arrange(grobs = grobs, nrow = 5)
@

@@

The categories we use in this study are operationally defined as follows:
\begin{description}
\item [Bowtie] Bowtie shapes are roughly quadrilateral, with two opposite concave faces. The remaining two faces can be convex or straight, and the concave faces may have straight portions, so long as there is a concave region. %Using this definition, shapes such as butterflies are included as bowties. \svp{I'm still on the fence - we can just not mention this, or we could move butterflies to "star"...} \mt{I think it makes a lot of sense to move them to star. That means redoing the poster/slides examples though... grrrrrr} I'm just commenting that bit out for now.
\item [Chevron] Chevron shapes include repeating parallel lines as well as individual ``v" shapes. They may be angular but can also be curved.
\item [Circle] Circles include ellipses and ovals; they must be round.
\item [Line] Lines are repeated and parallel; a more general definition of a line would be difficult to differentiate from many other patterns. Lines can be mildly curved.
\item [Polygon] Polygons are defined in this standard to have more than 4 sides. They include pentagons, hexagons, and octagons.
\item [Quadrilateral] Quadrilaterals (quads) have four sides. They may have rounded or square corners.
\item [Star] Stars are any shape with alternating concave and convex regions, or lines which emanate from a central point. ``X" and ``+" shapes are also classified as stars.
\item [Text] Text is any shape which would be identified as text by a reasonable human. In most cases, the text on the outsole images used is made up of Latin alphabet characters; the model will likely not recognize text in other scripts (but could be trained if non-Latin text images could be obtained).
\item [Triangle] Triangles are any three-sided figure. Like quadrilaterals, they can have rounded corners. In some cases, it is difficult to distinguish between a trapezoidal shape and a triangle when rounded corners are involved.
\item [Other] Other features which were marked include logos, various textures (including crepe, stippling, etc.), and smooth regions with no discernible features. These regions are grouped and provide additional information - that none of the previous nine categories are present.
\end{description}

@@
